---
title: 'MATH 216 - Notes: Visualization Customization'
author: "Emily Malcolm-White"
date: "03/11/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
---

# Load Packages, Import Data and Plot

We will be working with the `ToothGrowth` dataset which is available within the `tidyverse` package. It is data from an experiment on the effect of Vitamin C on Tooth Growth in Guinea Pigs. There are three variables: 

-  `len`  tooth length
-  `supp`  supplement type (VC or OJ)
-  `dose`  dose in mg (0.5,1,2)


```{r warning=FALSE, include=FALSE}
library(tidyverse)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```


```{r}
#print out 10 random rows from the dataset
ToothGrowth[runif(10,1,60),]
```


```{r}
#creates a side-by-side boxplot which we can customize
p <- ggplot(ToothGrowth, aes(x=dose, y=len)) + geom_boxplot()
p
```


# Continuous Axis Customizations

## Transformations

If the data is highly skew, you may be interested in transforming the data to get a better view. 

- `scale_x_log10()`, `scale_y_log10()` : for log10 transformation
- `scale_x_sqrt()`, `scale_y_sqrt()`: for sqrt transformation
- `scale_x_reverse()`, `scale_y_reverse()`: to reverse coordinates
- `coord_trans(x =“log10”, y=“log10”)` : possible values for x and y are “log2”, “log10”, “sqrt”, …
- `scale_x_continuous(trans=‘log2’)`, `scale_y_continuous(trans=‘log2’)` : another allowed value for the argument trans is ‘log10’


```{r}
p + scale_y_log10()
```

I personally like the using the above options because look at how the y-axis displays. It still displays the values for count, rather than displaying log count (which is hard to interpret). 

The other option would be to use `mutate` to log the data and then create a boxplot with that. Notice the axis displays log_lens, which is harder to interpret to a general audience. 

```{r}
ToothGrowth %>% 
  mutate(log_len = log(len)) %>% 
  ggplot(aes(x=dose, y=log_len)) + geom_boxplot()
```


## Change position of tick marks (continuous axis)

You may be interested in customizing where the tick marks go. You may want them closer together or farther apart.

```{r}
# Set tick marks on y axis
# a tick mark is shown on every 5
p + scale_y_continuous(breaks=seq(0,40,5))

# Tick marks can be spaced randomly
# I wouldn't recommend this, but you can do it!
p + scale_y_continuous(breaks=c(5,7.5, 20, 25))
```


## Changing formatting (dollar, scientific, comma, percent)

You may be interested in changing out the y-axis displays. For example, I noticed that some of your plots were displaying that values in scientific notation. If you want that, you can specific the label to be `scientific`. If you wish for it to display in the different way, you can use the label `comma`. 

If it's monetary value, you may be interested in the label `dollar` to add a "$" sign. If it's a percent, you may specific `percent` to add a "%" sign. 

```{r}
library(scales)
p + scale_y_continuous(labels = scientific)
```

```{r}
library(scales)
p + scale_y_continuous(labels = comma)
```

# Dates

Below is some code I used to create a dataset and graph (so we have something to practice customizing). 

```{r}
#Create time series data
df <- data.frame(
  date = seq(Sys.Date(), len=100, by="1 day")[sample(100, 50)],
  price = runif(50)
)
df <- df[order(df$date), ]


#Example plot
dp <- ggplot(df, aes(x=date, y=price)) + geom_line()
dp
```

You can customize how the date displays

```{r}
# Format: month(numeric)/day 
dp + scale_x_date(labels = date_format("%m/%d")) +
  theme(axis.text.x = element_text(angle=45))
# Format : week(numeric)
dp + scale_x_date(labels = date_format("%W"))
# Format: month(word) only
dp + scale_x_date(breaks = date_breaks("months"),
  labels = date_format("%b"))
```

We will spend some more time talking about dates when we get back from break... 

# Change the appearance text (font, face, colour, size, angle) 

Just so you know, you can totally customize any of the labels - the font, the color, the size, the angle, etc. 

`family` : font family
`face` : font face. Possible values are “plain”, “italic”, “bold” and “bold.italic”
`colour` : text color
`size` : text size in pts
`angle` : angle (in [0, 360])

For example, 

```{r}
p + theme(axis.text.x = element_text(face="bold", color="#993333", 
                           size=14, angle=45),
          axis.text.y = element_text(face="bold", color="#993333", 
                           size=14, angle=45))
```

For the record, I think this is totally ugly and doesn't improve the graph. BUT, it's possible. 

## Change the appearance of other elements 

You can also customize the title, subtitle, etc. in the same way. The code below may inspire you. 

```{r}
 # theme(plot.title=element_text(size=20, 
 #                                  face="bold", 
 #                                    family="American Typewriter",
 #                                    color="tomato",
 #                                    hjust=0.5,
 #                                    lineheight=1.2),  # title
 #            plot.subtitle=element_text(size=15, 
 #                                       family="American Typewriter",
 #                                       face="bold",
 #                                       hjust=0.5),  # subtitle
 #            plot.caption=element_text(size=15),  # caption
 #            axis.title.x=element_text(vjust=10,  
 #                                      size=15),  # X axis title
 #            axis.title.y=element_text(size=15),  # Y axis title
 #            axis.text.x=element_text(size=10, 
 #                                     angle = 30,
 #                                     vjust=.5),  # X axis text
 #            axis.text.y=element_text(size=10))  # Y axis text
```

