---
title: 'MATH 216 - Notes: Maps EXTRAS'
author: "Emily Malcolm-White"
output: 
  html_document:
    fig_crop: yes
    toc: true
    toc_float: true
    theme: yeti
---




# Using `maps`

The `maps` package contains several built in maps: world (for all countries in the world), france, italy, nz, usa, state (usa state boundaries), and county (usa counties). 

To reference each map you use `map_data("mapname")`. Rather than use `geom_sf` map data in this format needs to be mapped using `geom_polygon`. 

```{r}
world_map <- map_data("world")

ggplot(world_map, aes(long, lat)) + 
  geom_polygon(aes(group=group),fill="lightgrey", col="black") + 
  theme_light()
```


You can subset the data from particular country/countries.  

## Canada

```{r}
canada <- map_data("world", region = "canada")

ggplot(canada, aes(long, lat)) + 
  geom_polygon(aes(group=group), fill="darkgrey",col="black") + 
  theme_light() + 
  theme(legend.position = "none") 

#notice how this looks weird? This is because the longitude and latitude are not scaled properly... 
```

```{r}
canada <- map_data("world", region = "canada")

ggplot(canada, aes(long, lat)) + 
  geom_polygon(aes(group=group), fill="darkgrey",col="black") + 
  theme_light() + 
  theme(legend.position = "none") +
  coord_fixed(ratio = 1) 

#projection still doesn't look quite right... 
```



## EU

```{r}
some.eu.countries <- c(
  "Portugal", "Spain", "France", "Switzerland", "Germany",
  "Austria", "Belgium", "UK", "Netherlands",
  "Denmark", "Poland", "Italy", 
  "Croatia", "Slovenia", "Hungary", "Slovakia",
  "Czech republic"
)

europe <- map_data("world", region = some.eu.countries)

ggplot(europe, aes(long, lat)) + 
  geom_polygon(aes( group=group, fill=region), col="black") + 
  theme_light() + 
  theme(legend.position = "none") 

#this is an instance where you would want to re-visit the projection
```



## Adding Data from another dataset

```{r}
# Prepare the USArrests data
arrests <- USArrests 
arrests$region <- tolower(rownames(USArrests))
head(arrests)

# Retrieve the states map data and merge with crime data
states_map <- map_data("state")
arrests_map <- left_join(states_map, arrests, by = "region") #we will talk alot more about joins in the 

head(arrests_map)

# Create the map
ggplot(arrests_map, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = Murder), color = "black") +
  theme_light()
```

