---
title: 'MATH 216 - Notes: Visualization Customization'
author: "Emily Malcolm-White"
date: "03/11/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
---

# Load Packages, Import Data and Plot

```{r}
library(tidyverse)
# Tooth Growth Dataset is built in to tidyverse
# The Effect of Vitamin C on Tooth Growth in Guinea Pigs
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
head(ToothGrowth)
# len - tooth length
# supp - supplement type (VC or OJ)
# dose - dose in mg (0.5,1,2)
```


```{r}
p <- ggplot(ToothGrowth, aes(x=dose, y=len)) + geom_boxplot()
p
```


# Y-axis

## Transformations

`scale_x_log10()`, `scale_y_log10()` : for log10 transformation
`scale_x_sqrt()`, `scale_y_sqrt()`: for sqrt transformation
`scale_x_reverse()`, `scale_y_reverse()`: to reverse coordinates
`coord_trans(x =“log10”, y=“log10”)` : possible values for x and y are “log2”, “log10”, “sqrt”, …
`scale_x_continuous(trans=‘log2’)`, `scale_y_continuous(trans=‘log2’)` : another allowed value for the argument trans is ‘log10’


```{r}
p + scale_y_log10()
```

Better than using `mutate` (see the difference in the axes... )

```{r}
ToothGrowth %>% 
  mutate(log_len = log(len)) %>% 
  ggplot(aes(x=dose, y=log_len)) + geom_boxplot()
```


## Change position of tick marks (continuous axis)

```{r}
# Set tick marks on y axis
# a tick mark is shown on every 5
p + scale_y_continuous(breaks=seq(0,40,5))

# Tick marks can be spaced randomly
p + scale_y_continuous(breaks=c(5,7.5, 20, 25))
                     
# Remove tick mark labels and gridlines
p + scale_y_continuous(breaks=NULL)
```


## Changing formatting (dollar, scientific, comma, percent)

```{r}
library(scales)
p + scale_y_continuous(labels = scientific)

# other options are `dollar`, `percent`, `comma`
```

# X axis

## Change tick mark labels (discrete axis)

```{r}
# Solution 1
p + scale_x_discrete(breaks=c("0.5","1","2"),
        labels=c("Dose 0.5", "Dose 1", "Dose 2"))

# Solution 2 : same plot as solution 1
#p + scale_x_discrete(labels=c("0.5" = "Dose 0.5", "1" = "Dose 1", "2" = "Dose 2"))
```

## Change labels (dates)

```{r}

#Create time series data
df <- data.frame(
  date = seq(Sys.Date(), len=100, by="1 day")[sample(100, 50)],
  price = runif(50)
)
df <- df[order(df$date), ]


#Example plot
dp <- ggplot(df, aes(x=date, y=price)) + geom_line()
dp
```

```{r}
dp + scale_x_date(labels = date_format("%m/%d")) +
  theme(axis.text.x = element_text(angle=45))
# Format : Week
dp + scale_x_date(labels = date_format("%W"))
# Months only
dp + scale_x_date(breaks = date_breaks("months"),
  labels = date_format("%b"))
```

We will spend some more time talking about dates soon... 

# Change the appearance text (font, face, colour, size, angle) 

`family` : font family
`face` : font face. Possible values are “plain”, “italic”, “bold” and “bold.italic”
`colour` : text color
`size` : text size in pts
`angle` : angle (in [0, 360])

```{r}
p + theme(axis.text.x = element_text(face="bold", color="#993333", 
                           size=14, angle=45),
          axis.text.y = element_text(face="bold", color="#993333", 
                           size=14, angle=45))
```

# Change the appearance of other elements 

```{r}
 # theme(plot.title=element_text(size=20, 
 #                                  face="bold", 
 #                                    family="American Typewriter",
 #                                    color="tomato",
 #                                    hjust=0.5,
 #                                    lineheight=1.2),  # title
 #            plot.subtitle=element_text(size=15, 
 #                                       family="American Typewriter",
 #                                       face="bold",
 #                                       hjust=0.5),  # subtitle
 #            plot.caption=element_text(size=15),  # caption
 #            axis.title.x=element_text(vjust=10,  
 #                                      size=15),  # X axis title
 #            axis.title.y=element_text(size=15),  # Y axis title
 #            axis.text.x=element_text(size=10, 
 #                                     angle = 30,
 #                                     vjust=.5),  # X axis text
 #            axis.text.y=element_text(size=10))  # Y axis text
```

